---
title: "Reproducible Research: Storms and other weather events that cause both public health and economic problems in USA"
output: 
  html_document:
    keep_md: true
---

* *Author - Sanjeev Kumar Maheve*
* *Dated -  August 23, 2015*

### Synopsis
In this report we aim to explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. From these data, we found that Tornado causes major harmful injuries whereas Tornado, Heat, Lightening and Flood are major fatal weather events causes death in USA.

```{r setoptions,echo=TRUE}
library(knitr)
# Setting the global options.
opts_chunk$set(echo = TRUE)
```

### Reading in the dataset from year 1950 till end in November 2011
```{r cache=TRUE}
## Download the said file from the location given below
##
inputfile <- paste("storm_data", ".bz2", sep="")
if(!file.exists(inputfile)) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl, destfile = inputfile, method = "curl")
}

# Load the data as per default column type information.
df <- read.table(inputfile, 
                 header=TRUE, 
                 sep=",",
                 na.strings="",
                 strip.white = TRUE,
                 stringsAsFactors = FALSE)
str(df)
```

After reading in U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database between 1950 - 2011, there are ```902297``` observations in the dataset.

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r}
library(dplyr)
library(datasets)
library(ggplot2) 

# Group the data based on the event type and aggregate for 
# injuries and fatalities.
injuryStats <- aggregate (INJURIES~EVTYPE, data=df, FUN=sum, na.rm=TRUE) %>%
    filter(INJURIES > 0) %>%
    arrange(desc(INJURIES))
fatalStats <- aggregate (FATALITIES~EVTYPE, data=df, FUN=sum, na.rm=TRUE) %>%
    filter(FATALITIES > 0) %>%
    arrange(desc(FATALITIES))
```

#### Results (entire USA analysis) - showing health impacts due to weather events.
```{r}
# Lets show the first few observations in 2 columns plot. 
par(mfrow = c(1, 2))
par(mar=c(5.1,4.1,4.1,2.1))
barplot(injuryStats$INJURIES[1:10],
        names.arg = injuryStats$EVTYPE[1:10],
        main = "Top 10 harmful injuries",
        ylab = "Injuries",
        col="darkblue",
        cex.axis = 0.8,
        cex.names = 0.7,
        las = 2)
barplot(fatalStats$FATALITIES[1:10], 
        names.arg = fatalStats$EVTYPE[1:10],
        main = "Top 10 harmful fatalities",
        ylab = "Fatalities",
        col="red",
        cex.axis = 0.8,
        cex.names = 0.7,
        las = 2)
title("Health impact between 1950-2011 in USA", 
      outer = TRUE)
```

**Summary Report**
The weather event that was the most harmful to human health between 1950 - 2011 is **Tornado** followed by **TSTM Wind and Flood**. Refer the results below:

``` {r}
head(injuryStats)
head(fatalStats)
```

### Across the United States, which types of events have the greatest economic consequences?
The variables PROPDMGEXP and CROPDMGEXP have the factor of multiplicity of the variables PROPDMG and CROPDMG with the values:

* H -> 100  
* K -> 1000  
* M -> 1000,000  
* B -> 1000,000,000  

```{r}
# Map all the unit mentioned above to common base unit.
# PROP
df$PROPDMGEXP[is.na(df$PROPDMGEXP)]<-0
df$PROPDMGEXP <- gsub("H", 1e2, df$PROPDMGEXP, ignore.case = TRUE)
df$PROPDMGEXP <- gsub("K", 1e3, df$PROPDMGEXP, ignore.case = TRUE)
df$PROPDMGEXP <- gsub("M", 1e6, df$PROPDMGEXP, ignore.case = TRUE)
df$PROPDMGEXP <- gsub("B", 1e9, df$PROPDMGEXP, ignore.case = TRUE)
df$PROPDMGEXP <- as.numeric(df$PROPDMGEXP)
# CROP
df$CROPDMGEXP[is.na(df$CROPDMGEXP)]<-0
df$CROPDMGEXP <- gsub("H", 1e2, df$CROPDMGEXP, ignore.case = TRUE)
df$CROPDMGEXP <- gsub("K", 1e3, df$CROPDMGEXP, ignore.case = TRUE)
df$CROPDMGEXP <- gsub("M", 1e6, df$CROPDMGEXP, ignore.case = TRUE)
df$CROPDMGEXP <- gsub("B", 1e9, df$CROPDMGEXP, ignore.case = TRUE)
df$CROPDMGEXP <- as.numeric(df$CROPDMGEXP)
# Create new column(s) using combination of existing columns for 
# PROP and CROP
df <- df %>% 
    mutate(PROPTOTALDMG = PROPDMG * PROPDMGEXP) %>%
    mutate(CROPTOTALDMG = CROPDMG * CROPDMGEXP)
df <- df %>% mutate(ECONOMYTOTALDMG = PROPTOTALDMG + CROPTOTALDMG)
str(df)

# Group the data based on the event type and aggregate for 
# property, crop and combined for economy.
propStats <- aggregate (PROPTOTALDMG~EVTYPE, data=df, FUN=sum, na.rm=TRUE) %>%
    filter(PROPTOTALDMG > 0) %>%
    arrange(desc(PROPTOTALDMG))
cropStats <- aggregate (CROPTOTALDMG~EVTYPE, data=df, FUN=sum, na.rm=TRUE) %>%
    filter(CROPTOTALDMG > 0) %>%
    arrange(desc(CROPTOTALDMG))
economyStats <- aggregate (ECONOMYTOTALDMG~EVTYPE, data=df, FUN=sum, na.rm=TRUE) %>%
    filter(ECONOMYTOTALDMG > 0) %>%
    arrange(desc(ECONOMYTOTALDMG))
```

#### Results (entire USA analysis) - showing economy consequencies due to weather events.
```{r}
# Lets show the first few observations in 2 columns plot.
barplot(economyStats$ECONOMYTOTALDMG[1:10],
        names.arg = economyStats$EVTYPE[1:10],
        main = "Economy Damage (USD)",
        ylab = "Economy (USD)",
        col="red",
        cex.axis = 0.8,
        cex.names = 0.7,
        las = 2)
par(mfrow = c(1, 2))
par(mar=c(5.1,4.1,4.1,2.1))
barplot(propStats$PROPTOTALDMG[1:10],
        names.arg = propStats$EVTYPE[1:10],
        main = "Property Damage (USD)",
        ylab = "Property (USD)",
        col="magenta",
        cex.axis = 0.8,
        cex.names = 0.7,
        las = 2)
barplot(cropStats$CROPTOTALDMG[1:10], 
        names.arg = cropStats$EVTYPE[1:10],
        main = "Crop Damage (USD)",
        ylab = "Crop (USD)",
        col="skyblue",
        cex.axis = 0.8,
        cex.names = 0.7,
        las = 2)
title("Economy consequence between 1950-2011 in USA", 
      outer = TRUE)
```

**Summary Report**  
The weather event that resulted in bad economy consequences between 1950 - 2011 is **Flood** followed by **Hurricane/Typhoon and Tornado** for property and **Drought** followed by **Flood** for Crops.  
``` {r}
head(economyStats)
head(propStats)
head(cropStats)
```
